Identifying Factors That Affect Political Freedom Within the OECD

### 2020-11-29 paper/paper.Rmd
meeting

2020 is an election year, which has many of use reflecting on political freedoms. Political freedoms are important because they have a dramatic impact on the quality of life for the citizens of a country. Such reflection has us asking: What factors are associated with an increase in political freedom around the world? Political freedom is an abstract concept that does not lend well to quantification. However, Freedomhouse [@Freedomhouse], a U.S. based think tank and research institute, attempts this by assigning an index of political rights to each country every year based on a fixed criteria. Using the Papaja Package by [@R-papaja] we take Rmarkdown directly to APA format.

The **Political Rights** Index is comprised of three subcategories. Electoral process, political participation, and functioning government. Electoral process is a score based on how the current government leaders were selected. How fairly the positions were obtained and kept. As well as the fairness of the current electoral laws, how they are implemented and the degree to which there is an independent judiciary. Political Participation is evaluated on four criterion, right to form parties, realistic opposition to current power, political choice free from military, religious powers, economic oligarchies, or any other unaccountable body, and various minority groups having full political rights. Functioning government, measures the level of autonomy of the current heads of government when determining policies, safeguards against corruption, and transparency of government operations. Freedom house has measured political freedom in this manner for every country from the years 1978 to present.

Although such a metric is far from perfect and often debatable, this will serve as a reasonable measure of political freedom around the world to study political freedom. We obtained the dataset from the Gapminder Foundation [@Gapminder], a non-profit organization that studies and promotes economic development. After carefully selecting a dependent variable, we browsed data that were also available from Gapminder and selected variables that we felt would be helpful in explaining political freedom in various countries. To explain a country's political rights index we selected: **Corruption Perception Index**, **Education Expenditure**, **Electricity Use**, **Gini Coefficient**, **Internet Users** as a percentage, **Labor Force Participation** Rate, **Military Spending** as percentage of GDP, and **Murders** per million.

@@

These data sets were separate so we downloaded them individually, cleaned and combined them together using R [@R-base] and the Tidyverse Package [@R-tidyverse]. This resulted in a data set with all 196 countries from years 1950 to 2030 (data after 2019 were projections). It was immediately apparent that there was a substantial amount of missing data in an obvious pattern. Less developed countries tended to have more missing data points, and the further back in time the more missing data. Had we conducted this study as it was it would have been severely biased towards more developed countries as missing data points would be dropped. This forced us to narrow the scope of the study and focus only on countries within the Organization for Economic Co-operation and Development (OECD). This subset of countries has reasonably complete data. Therefore, we removed data from countries that are not members of the OECD and selected only the years 2000 through 2018.

@@

We examined the results of that we did not find compelling evidence to support making a such a transformation and the data are roughly linear. We will not be making any non-linear transformation or variables.  With a dataset that contains flaws but still reasonably holds assumptions we performed the best subsection algorithm to obtain a model with the best fit. This lead us to choose between the following models. From the graph of the subsets we can see that the increase in $R^{2}_a$ and $AIC$ by adding additional parameters diminishes quite rapidly.

@@

From the best subsets, it was appropriate to narrow our model choices down to three. First, the model with six degrees of freedom, next one with five, and one with four. The results from these three models is in Table 2 made using Stargazer [@R-stargazer]. The model with six degrees of freedom contains **Internet Users** which is 0.001, which is not statistically significant and even if it was a one percent increase in the percentage of people using the internet associated with a 0.001 increase in political freedom is not politically significant in any meaningful way. Looking at the model with five degrees of freedom, all of our coefficients are statistically significant at $p < 0.01$. The model with four degrees of freedom also has statistically significant coefficients and we would pick it had life expectancy not been significant. Counterintuitively, politically freedom seems to decrease as life expectancy increases. We are not qualified to attempt explain that we can only speculate. We are going to choose the model with five degrees of freedom. So we arrived at the model with five degrees of freedom.

@@

(a + b) / (c + d + e) 
```
The scatter plots look a bit strange at first because of the nature of the discrete response variable. However, this does seem to fit the data reasonably. There is a fairly linear relationship with a clear sloping line which, accept for life expectancy, concur with the regression coefficient. Life expectancy on its own has a positive association with political freedom, however when holding the other covariates constant we find that slope to be negative.

@@

Now we will examine some remedial measures to analyze the performance and identify problems. 
Let's start with plots of residuals. Figure 2, contains a histogram of the residuals adjacent to a QQ normal plot. It is clear that there are substantial outliers at the lower end of political freedom. Upon further inspection we found that these outliers are Turkey from 2016 to 2018. These residuals are large and suggest a substantial loss in political freedom. This is confirmed by Freedomhouse [@Freedomhouse] reporting that a coup attempt in Turkey resulted in a political retaliation against perceived opponents and constitutional changes were made that concentrated political power to the president. After much discussion we concluded that since the regression results were not severely affected and this is not an entry error there is not a compelling reason to eliminate it.  
It was clear to us that there is substantial error variance and likely not constant from the positive residuals to negative residuals. A Brown-Forsythe test confirmed our suspicions telling us that the error variance is significantly higher when the residuals are negative. This is likely due to the fact that our sample includes predominately values that are on the upper end of the political freedom.

>>>>

2020 is an election year, which has many of us reflecting on political freedom. Political freedoms are important because they have a dramatic impact on the quality of life for the citizens of a country. Such reflection has us asking: What factors are associated with an increase in political freedom around the world? Political freedom is an abstract concept that does not lend well to quantification. However, Freedomhouse [@Freedomhouse], a U.S. based think tank and research institute, attempts this by assigning an index of political rights to each country every year based on a fixed criteria. Using the Papaja Package by [@R-papaja] we take Rmarkdown directly to APA format.

The **Political Rights** Index is comprised of three subcategories. Electoral process, political participation, and functioning government. Electoral process is a score based on how the current government leaders were selected, mainly how fairly the positions were obtained and kept. As well as the fairness of the current electoral laws, how they are implemented and the degree to which there is an independent judiciary. Political Participation is evaluated on four criterion, right to form parties, realistic opposition to current power, political choice free from military, religious powers, economic oligarchies, or any other unaccountable body, and various minority groups having full political rights. Functioning government, measures the level of autonomy of the current heads of government when determining policies, safeguards against corruption, and transparency of government operations. Freedom house has measured political freedom in this manner for every country from the years 1978 to present.

Although such a metric is imperfect and debatable, this will serve as a reasonable measure of political freedom around the world. We obtained the dataset from the Gapminder Foundation [@Gapminder], a non-profit organization that studies and promotes economic development. After carefully selecting a dependent variable, we browsed data that were also available from Gapminder and selected variables that we felt would be helpful in explaining political freedom in various countries. To explain a country's political rights index we selected: **Corruption Perception Index**, **Education Expenditure**, **Electricity Use**, **Gini Coefficient** a measure of economic inequality, **Internet Users** as a percentage, **Labor Force Participation** Rate, **Military Spending** as percentage of GDP, and **Murders** per million.

@@

These data sets were separate so we downloaded them individually, cleaned and combined them together using R [@R-base] and the Tidyverse Package [@R-tidyverse]. This resulted in a data set with all 196 countries from years 1950 to 2030 (data after 2019 were projections). It was immediately apparent that there was a substantial amount of missing data in an obvious pattern. Less developed countries tended to have more missing data points, and the further back in time the more missing data. Had we conducted this study as it was it would have been severely biased towards more developed countries as missing data points would be dropped. This forced us to narrow the scope of the study and focus only on countries within the Organization for Economic Co-operation and Development (OECD). This subset of countries has much more complete data. Therefore, we removed data from countries that are not members of the OECD and selected only the years 2000 through 2018.

@@

We examined the results of that we did not find compelling evidence to support making a such a transformation and the data are roughly linear. We will not be making any non-linear transformation of variables.  With a dataset that contains flaws but still reasonably holds assumptions we performed the best subset algorithm to obtain a model with the best fit. From the graph of the subsets we can see that the increase in $R^{2}_a$ and $AIC$ by adding additional parameters diminishes quite rapidly.

@@

From the best subsets, it was appropriate to narrow our model choices down to three. First, the model with six degrees of freedom, next one with five, and one with four. The results from these three models is in Table 2 made using Stargazer [@R-stargazer]. The model with six degrees of freedom contains **Internet Users** which is 0.003, which is statistically significant, however, a one percent increase in the percentage of people using the internet is associated with a 0.003 increase in political freedom is not politically significant. Looking at the model with five degrees of freedom, all of our coefficients are significant at the 99% level. The model with four degrees of freedom also has statistically significant coefficients and we would pick it had life expectancy not been significant. Counterintuitively, politically freedom seems to decrease as life expectancy increases. We are not qualified to explain that we can only speculate. This is how we arrived at the model with five degrees of freedom.

@@

(a + b) / (c + d + e) 
```
The scatter plots look a bit strange at first because of the nature of the discrete response variable. However, this does seem to fit the data reasonably. There is a fairly linear relationship with a clear sloping line which, except for life expectancy, concur with the regression coefficient. Life expectancy on its own has a positive association with political freedom, however when holding the other covariates constant we find that slope to be negative.

@@

Now we will examine some regression diagnostics to analyze the performance and identify problems. Let's start with plots of residuals. Figure 2, contains a histogram of the residuals adjacent to a QQ normal plot. It is clear that there are substantial outliers at the lower end of political freedom. Upon further inspection we found that these outliers are Turkey from 2016 to 2018. These residuals are large and suggest a substantial loss in political freedom. This is confirmed by Freedomhouse [@Freedomhouse] reporting that a coup attempt in Turkey resulted in a political retaliation against perceived opponents and constitutional changes were made that concentrated political power to the president. After much discussion we concluded that since the regression results were not severely affected and this is not an entry error there is not a compelling reason to eliminate it.  
It was clear to us that there is substantial error variance and likely not constant from the positive residuals to negative residuals. A Brown-Forsythe test confirmed our suspicions telling us that the error variance is significantly higher when the residuals are negative. This is likely due to the fact that our sample includes predominately values that are on the upper end of the political freedom.

### 2020-11-29 paper/paper.tex
meeting

2020 is an election year, which has many of use reflecting on political freedoms. Political freedoms are important because they have a dramatic impact on the quality of life for the citizens of a country. Such reflection has us asking: What factors are associated with an increase in political freedom around the world? Political freedom is an abstract concept that does not lend well to quantification. However, Freedomhouse (\emph{Countries and Territories}, 2020), a U.S. based think tank and research institute, attempts this by assigning an index of political rights to each country every year based on a fixed criteria. Using the Papaja Package by (Aust \& Barth, 2020) we take Rmarkdown directly to APA format.

The \textbf{Political Rights} Index is comprised of three subcategories. Electoral process, political participation, and functioning government. Electoral process is a score based on how the current government leaders were selected. How fairly the positions were obtained and kept. As well as the fairness of the current electoral laws, how they are implemented and the degree to which there is an independent judiciary. Political Participation is evaluated on four criterion, right to form parties, realistic opposition to current power, political choice free from military, religious powers, economic oligarchies, or any other unaccountable body, and various minority groups having full political rights. Functioning government, measures the level of autonomy of the current heads of government when determining policies, safeguards against corruption, and transparency of government operations. Freedom house has measured political freedom in this manner for every country from the years 1978 to present.

Although such a metric is far from perfect and often debatable, this will serve as a reasonable measure of political freedom around the world to study political freedom. We obtained the dataset from the Gapminder Foundation (Rosling, 2020), a non-profit organization that studies and promotes economic development. After carefully selecting a dependent variable, we browsed data that were also available from Gapminder and selected variables that we felt would be helpful in explaining political freedom in various countries. To explain a country's political rights index we selected: \textbf{Corruption Perception Index}, \textbf{Education Expenditure}, \textbf{Electricity Use}, \textbf{Gini Coefficient}, \textbf{Internet Users} as a percentage, \textbf{Labor Force Participation} Rate, \textbf{Military Spending} as percentage of GDP, and \textbf{Murders} per million.

@@

These data sets were separate so we downloaded them individually, cleaned and combined them together using R (R Core Team, 2020) and the Tidyverse Package (Wickham et al., 2019). This resulted in a data set with all 196 countries from years 1950 to 2030 (data after 2019 were projections). It was immediately apparent that there was a substantial amount of missing data in an obvious pattern. Less developed countries tended to have more missing data points, and the further back in time the more missing data. Had we conducted this study as it was it would have been severely biased towards more developed countries as missing data points would be dropped. This forced us to narrow the scope of the study and focus only on countries within the Organization for Economic Co-operation and Development (OECD). This subset of countries has reasonably complete data. Therefore, we removed data from countries that are not members of the OECD and selected only the years 2000 through 2018.

@@

We examined the results of that we did not find compelling evidence to support making a such a transformation and the data are roughly linear. We will not be making any non-linear transformation or variables. With a dataset that contains flaws but still reasonably holds assumptions we performed the best subsection algorithm to obtain a model with the best fit. This lead us to choose between the following models. From the graph of the subsets we can see that the increase in \(R^{2}_a\) and \(AIC\) by adding additional parameters diminishes quite rapidly.

@@

From the best subsets, it was appropriate to narrow our model choices down to three. First, the model with six degrees of freedom, next one with five, and one with four. The results from these three models is in Table 2 made using Stargazer (Hlavac, 2018). The model with six degrees of freedom contains \textbf{Internet Users} which is 0.001, which is not statistically significant and even if it was a one percent increase in the percentage of people using the internet associated with a 0.001 increase in political freedom is not politically significant in any meaningful way. Looking at the model with five degrees of freedom, all of our coefficients are statistically significant at \(p < 0.01\). The model with four degrees of freedom also has statistically significant coefficients and we would pick it had life expectancy not been significant. Counterintuitively, politically freedom seems to decrease as life expectancy increases. We are not qualified to attempt explain that we can only speculate. We are going to choose the model with five degrees of freedom. So we arrived at the model with five degrees of freedom.

\includegraphics{paper_files/figure-latex/unnamed-chunk-5-1.pdf}
The scatter plots look a bit strange at first because of the nature of the discrete response variable. However, this does seem to fit the data reasonably. There is a fairly linear relationship with a clear sloping line which, accept for life expectancy, concur with the regression coefficient. Life expectancy on its own has a positive association with political freedom, however when holding the other covariates constant we find that slope to be negative.

@@

Now we will examine some remedial measures to analyze the performance and identify problems.
Let's start with plots of residuals. Figure 2, contains a histogram of the residuals adjacent to a QQ normal plot. It is clear that there are substantial outliers at the lower end of political freedom. Upon further inspection we found that these outliers are Turkey from 2016 to 2018. These residuals are large and suggest a substantial loss in political freedom. This is confirmed by Freedomhouse (\emph{Countries and Territories}, 2020) reporting that a coup attempt in Turkey resulted in a political retaliation against perceived opponents and constitutional changes were made that concentrated political power to the president. After much discussion we concluded that since the regression results were not severely affected and this is not an entry error there is not a compelling reason to eliminate it.\\
It was clear to us that there is substantial error variance and likely not constant from the positive residuals to negative residuals. A Brown-Forsythe test confirmed our suspicions telling us that the error variance is significantly higher when the residuals are negative. This is likely due to the fact that our sample includes predominately values that are on the upper end of the political freedom.

>>>>

2020 is an election year, which has many of us reflecting on political freedom. Political freedoms are important because they have a dramatic impact on the quality of life for the citizens of a country. Such reflection has us asking: What factors are associated with an increase in political freedom around the world? Political freedom is an abstract concept that does not lend well to quantification. However, Freedomhouse (\emph{Countries and Territories}, 2020), a U.S. based think tank and research institute, attempts this by assigning an index of political rights to each country every year based on a fixed criteria. Using the Papaja Package by (Aust \& Barth, 2020) we take Rmarkdown directly to APA format.

The \textbf{Political Rights} Index is comprised of three subcategories. Electoral process, political participation, and functioning government. Electoral process is a score based on how the current government leaders were selected, mainly how fairly the positions were obtained and kept. As well as the fairness of the current electoral laws, how they are implemented and the degree to which there is an independent judiciary. Political Participation is evaluated on four criterion, right to form parties, realistic opposition to current power, political choice free from military, religious powers, economic oligarchies, or any other unaccountable body, and various minority groups having full political rights. Functioning government, measures the level of autonomy of the current heads of government when determining policies, safeguards against corruption, and transparency of government operations. Freedom house has measured political freedom in this manner for every country from the years 1978 to present.

Although such a metric is imperfect and debatable, this will serve as a reasonable measure of political freedom around the world. We obtained the dataset from the Gapminder Foundation (Rosling, 2020), a non-profit organization that studies and promotes economic development. After carefully selecting a dependent variable, we browsed data that were also available from Gapminder and selected variables that we felt would be helpful in explaining political freedom in various countries. To explain a country's political rights index we selected: \textbf{Corruption Perception Index}, \textbf{Education Expenditure}, \textbf{Electricity Use}, \textbf{Gini Coefficient} a measure of economic inequality, \textbf{Internet Users} as a percentage, \textbf{Labor Force Participation} Rate, \textbf{Military Spending} as percentage of GDP, and \textbf{Murders} per million.

@@

These data sets were separate so we downloaded them individually, cleaned and combined them together using R (R Core Team, 2020) and the Tidyverse Package (Wickham et al., 2019). This resulted in a data set with all 196 countries from years 1950 to 2030 (data after 2019 were projections). It was immediately apparent that there was a substantial amount of missing data in an obvious pattern. Less developed countries tended to have more missing data points, and the further back in time the more missing data. Had we conducted this study as it was it would have been severely biased towards more developed countries as missing data points would be dropped. This forced us to narrow the scope of the study and focus only on countries within the Organization for Economic Co-operation and Development (OECD). This subset of countries has much more complete data. Therefore, we removed data from countries that are not members of the OECD and selected only the years 2000 through 2018.

@@

We examined the results of that we did not find compelling evidence to support making a such a transformation and the data are roughly linear. We will not be making any non-linear transformation of variables. With a dataset that contains flaws but still reasonably holds assumptions we performed the best subset algorithm to obtain a model with the best fit. From the graph of the subsets we can see that the increase in \(R^{2}_a\) and \(AIC\) by adding additional parameters diminishes quite rapidly.

@@

From the best subsets, it was appropriate to narrow our model choices down to three. First, the model with six degrees of freedom, next one with five, and one with four. The results from these three models is in Table 2 made using Stargazer (Hlavac, 2018). The model with six degrees of freedom contains \textbf{Internet Users} which is 0.003, which is statistically significant, however, a one percent increase in the percentage of people using the internet is associated with a 0.003 increase in political freedom is not politically significant. Looking at the model with five degrees of freedom, all of our coefficients are significant at the 99\% level. The model with four degrees of freedom also has statistically significant coefficients and we would pick it had life expectancy not been significant. Counterintuitively, politically freedom seems to decrease as life expectancy increases. We are not qualified to explain that we can only speculate. This is how we arrived at the model with five degrees of freedom.

\includegraphics{paper_files/figure-latex/unnamed-chunk-5-1.pdf}
The scatter plots look a bit strange at first because of the nature of the discrete response variable. However, this does seem to fit the data reasonably. There is a fairly linear relationship with a clear sloping line which, except for life expectancy, concur with the regression coefficient. Life expectancy on its own has a positive association with political freedom, however when holding the other covariates constant we find that slope to be negative.

@@

Now we will examine some regression diagnostics to analyze the performance and identify problems. Let's start with plots of residuals. Figure 2, contains a histogram of the residuals adjacent to a QQ normal plot. It is clear that there are substantial outliers at the lower end of political freedom. Upon further inspection we found that these outliers are Turkey from 2016 to 2018. These residuals are large and suggest a substantial loss in political freedom. This is confirmed by Freedomhouse (\emph{Countries and Territories}, 2020) reporting that a coup attempt in Turkey resulted in a political retaliation against perceived opponents and constitutional changes were made that concentrated political power to the president. After much discussion we concluded that since the regression results were not severely affected and this is not an entry error there is not a compelling reason to eliminate it.\\
It was clear to us that there is substantial error variance and likely not constant from the positive residuals to negative residuals. A Brown-Forsythe test confirmed our suspicions telling us that the error variance is significantly higher when the residuals are negative. This is likely due to the fact that our sample includes predominately values that are on the upper end of the political freedom.
